### 2012-02-24 Krantz-v3/Chapters/chapter1/model.tex
Added booke chapter.

HSLDA is a model for hierarchically, multiply-labeled, bag-of-word data.  We
will refer to individual groups of bag-of-word data as documents.  Let $w_{n,d}
\in \Sigma$ be the $n$th observation in the $d$th document.  Let $\mathbf{w}_d
= \{w_{1,d},\ldots,w_{1,N_d}\}$ be the  set of $N_d$ observations in document
$d$.  Let there be $D$ such documents and let the size of the vocabulary be
$V=|\Sigma|$.  Let the set of labels be $\mathcal{L}=\left\{
  l_{1},l_{2},\ldots,l_{\left|\mathcal{L}\right|}\right\} $. Each label
$l \in \mathcal{L}$, except the root, has a parent $\mathrm{pa}(l) \in \mathcal{L}$
also in the set of labels. 
 We will for exposition purposes assume that this label set has hard ``is-a''
 parent-child constraints (explained later), although this assumption can be
 relaxed at the cost of more computationally complex inference.  Such a label hierarchy forms a multiply rooted tree.  Without loss of generality we will consider a tree with a single root $r\in\mathcal{L}$.  Each document has a variable $y_{l,d} \in \{-1,1\}$ for every label which indicates whether the label is applied to document $d$ or not.   In most cases $y_{i,d}$ will be unobserved, in some cases we will be able to fix its value because of  constraints on the label hierarchy, and in the relatively minor remainder its value will be observed.  In the applications we consider, only positive labels are observed.

@@

% generalize to arbitrary codes (not just ICD-9's
%
\begin{figure}[t]
%tbp] %  figure placement: here, top, bottom, or page
 \centering \includegraphics[scale=0.4]{Graphical_Model-final} \caption{HSLDA graphical model}

@@

%In HSLDA, the bag-of-word document data is modeled using the LDA
%mixed-membership mixture model with global topic estimation.
In HSLDA, documents are modeled using the LDA mixed-membership mixture model
with global topic estimation. Label responses are generated using a conditional
hierarchy of probit regressors. The HSLDA graphical model is given in
Figure~\ref{fig:graphical_model}. In the model, $K$ is the number of LDA
``topics'' (distributions over the elements of $\Sigma$), $\boldsymbol\phi_k$
is a distribution over ``words,'' $\boldsymbol\theta_d$ is a document-specific
distribution over topics, $\boldsymbol\beta$ is a global distribution over
topics, Dir$_{K}(\cdot)$ is a $K$-dimensional Dirichlet distribution,
$\mathcal{N}_{K}(\cdot)$ is the $K$-dimensional Normal distribution,
$\mathbf{I}_{K}$ is the $K$ dimensional identity matrix,  $\mathbf{1}_d$ is the
$d$-dimensional vector of all ones, and $\mathbb{I}(\cdot)$ is an indicator
function that takes the value $1$ if its argument is true and $0$ otherwise.
The following procedure describes how to generate from the HSLDA generative
model.

@@

%We will now describe the stochastic generative process which defines
%our model. The 
%Given the number of topics, K, and broad gamma priors on hyperparameters,
%the generative process is as follows:

\begin{enumerate}
\item For each topic $k=1,\ldots,K$

@@

\begin{itemize}
\item Draw a label application coefficient $\boldsymbol\eta_{l}\mid\mu,\sigma\sim\mathcal{N}_{K}(\mu \mathbf{1}_K,\sigma \mathbf{I}_{K})$  
\end{itemize}
\item Draw the global topic proportions $\boldsymbol\beta\mid\alpha'\sim{\rm Dir}_{K}\left(\alpha^{\prime}\mathbf{1}_K\right)$
\item For each document $d=1,\ldots,D$

@@

\begin{itemize}
\item Draw $a_{l,d}\mid \bar{\mathbf{z}}_d,\boldsymbol\eta_{l},y_{\mathrm{pa}(l),d}\sim\begin{cases}
\mathcal{N}(\bar{\mathbf{z}}^{T}_d\boldsymbol\eta_{l},1), & y_{\mathrm{pa}(l),d}=1\\
\mathcal{N}(\bar{\mathbf{z}}^{T}_d\boldsymbol\eta_{l},1)\mathbb{I}(a_{l,d}<0), & y_{\mathrm{pa}(l),d}=-1\end{cases}$ %\item Draw $a_{l, d} \ | \ z_{1:N_d,d}, \boldsymbol\beta_l \sim \mathcal{N} \left(\bar z_d^{T} \boldsymbol\beta_{l},1\right)$, where $\bar z_d=N_d^{-1}\sum_{n=1}^{N_d}z_{n,d}$

\item Apply label $l$ to document $d$ according to $a_{l,d}$ \[\hspace{-2cm}
y_{l,d}\mid a_{l,d}=\begin{cases}
\ \ \ 1 & \text{if \ensuremath{a_{l,d}>0}} \\% and \ensuremath{y_{{\rm \mathrm{pa}}(l),d}=1}}\\
-1 & \text{otherwise}\end{cases}\]

\end{itemize}
\end{itemize}
\end{enumerate}

@@

The second half of step 4 is a substantial part of our contribution to the
general class of supervised LDA models.  Here, each document is labeled generatively using a hierarchy of conditionally dependent probit regressors \cite{gelmanbda04}.  For every label $l \in \mathcal{L}$, both the empirical topic distribution for document $d$ and whether or not its parent label was applied (i.e.~$\mathbb{I}(y_{\mathrm{pa}(l),d}=1)$) are used to determine whether or not  label $l$ is to be applied to  document $d$ as well.  Note that label $y_{l,d}$ can only be applied to document $d$ if its parent label $\mathrm{pa}(l)$ is also applied (these expressions are specific to is-a constraints but can be modified to accommodate different constraints).  The regression coefficients $\boldsymbol\eta_l$ are independent a priori, however, the hierarchical coupling in this model induces a posteriori dependence.   The net effect of this is that label predictors deeper in the label hierarchy are able to focus on finding specific, conditional labeling features.  We believe this to be a significant source of the empirical label prediction improvement we observe experimentally.  We test this hypothesis in Section~\ref{sec:experiments}.

Note that the choice of variables $a_{l,d}$ and how they are distributed were  driven at least in part by posterior inference efficiency considerations.  In particular, choosing probit-style auxiliary variable distributions for the $a_{l,d}$'s  yields conditional posterior distributions for both the auxiliary variables \eqref{eqn:a_l_d} and the regression coefficients \eqref{eqn:regression_param_conditional} which are analytic.  This simplifies posterior inference substantially.

@@

In the common case where no negative labels are observed (like the example applications we consider in Section~\ref{sec:experiments}), the model must be explicitly biased towards generating data that has negative labels in order to keep it from learning to assign all labels to all documents.   This is a common problem in modeling unbalanced data.  To see how this model can be biased in this way we draw the reader's attention to the $\mu$ parameter and, to a lesser extent, the $\sigma$ parameter above.  Because $\bar{\mathbf{z}}_d$ is always positive, setting $\mu$ to a negative value results in a bias towards negative labelings, i.e.~for large negative values of $\mu$, all labels become a priori more likely to be negative ($y_{l,d}=-1$).  We explore the ability of $\mu$ to bias out-of-sample label prediction performance in Section~\ref{sec:experiments}.

>>>>

HSLDA is a model for hierarchically, multiply-labeled, bag-of-word data.  We
call individual groups of bag-of-word data documents (although they could be any data represented by bags of features).  Let $w_{n,d}
\in \Sigma$ be the $n$th observation in the $d$th document.  Let $\mathbf{w}_d
= \{w_{1,d},\ldots,w_{1,N_d}\}$ be the  set of $N_d$ observations in document
$d$.  Let there be $D$ such documents and let the size of the vocabulary be
$V=|\Sigma|$.  Let the set of labels be $\mathcal{L}=\left\{
  l_{1},l_{2},\ldots,l_{\left|\mathcal{L}\right|}\right\} $. Each label
$l \in \mathcal{L}$, except the root, has a parent $\mathrm{pa}(l) \in \mathcal{L}$
also in the set of labels. 
 We will for exposition purposes assume that this label set has hard ``is-a''
 parent-child constraints (explained later), although this assumption can be
 relaxed at the cost of more computationally complex inference.  Such a label hierarchy forms a multiply rooted tree.  Without loss of generality we will consider a tree with a single root $r\in\mathcal{L}$.  Each document has a variable $y_{l,d} \in \{-1,1\}$ for every label which indicates whether the label is applied to document $d$ or not.   In most cases $y_{i,d}$ will be unobserved, in some cases we will be able to fix its value because of  constraints on the label hierarchy, and in the relatively minor remainder its value will be observed.  In the applications we demonstrate, only positive labels are observed.   This may not be true of all applications; however, positive-only label imbalance is a common problem.  How we solve this problem will be discussed later.

@@

% generalize to arbitrary codes (not just ICD-9's
%
\begin{figure}[t]
%tbp] %  figure placement: here, top, bottom, or page
 \centering \includegraphics[scale=0.4]{Chapters/chapter1/Graphical_Model-final} \caption{Hierarchically supervised latent Dirichlet allocation (HSLDA) graphical model.}

@@

In HSLDA, the bag-of-word document data is modeled using LDA with
 full, hierarchical topic estimation (i.e. with the global topic estimated too rather than assumed).
%In HSLDA, documents are modeled using the LDA mixed-membership mixture model
%with global topic estimation. 
Label responses are modeled using a conditional
hierarchy of probit regressors. The HSLDA graphical model is given in
Figure~\ref{fig:graphical_model}.

 In the corresponding mathematical generative description of HSLDA to follow, $K$ is the number of LDA
``topics'' (distributions over the elements of $\Sigma$), $\boldsymbol\phi_k$
is a distribution over ``words,'' $\boldsymbol\theta_d$ is a document-specific
distribution over topics, $\boldsymbol\beta$ is a global distribution over
topics, Dir$_{K}(\cdot)$ is a $K$-dimensional Dirichlet distribution,
$\mathcal{N}_{K}(\cdot)$ is the $K$-dimensional Normal distribution,
$\mathbf{I}_{K}$ is the $K$ dimensional identity matrix,  $\mathbf{1}_d$ is the
$d$-dimensional vector of all ones, and $\mathbb{I}(\cdot)$ is an indicator
function that takes the value $1$ if its argument is true and $0$ otherwise.

\subsection{Ancestral sampling for HSLDA}

The following procedure describes how to generate from the HSLDA generative
model.

@@

%We will now describe the stochastic generative process which defines
%our model. The 
%Given the number of topics, K, and broad gamma priors on hyperparameters,
%the generative process is as follows: 
\begin{shortbox}
\Boxhead{HSLDA Ancestral Sampler}
\begin{enumerate}
\item For each topic $k=1,\ldots,K$

@@

\begin{itemize}
\item Draw a label weight vector $\boldsymbol\eta_{l}\mid\mu,\sigma\sim\mathcal{N}_{K}(\mu \mathbf{1}_K,\sigma \mathbf{I}_{K})$  
\end{itemize}
\item Draw the global topic proportions $\boldsymbol\beta\mid\alpha'\sim{\rm Dir}_{K}\left(\alpha^{\prime}\mathbf{1}_K\right)$
\item For each document $d=1,\ldots,D$

@@

\begin{itemize}
\item Draw 
\begin{eqnarray}\lefteqn{a_{l,d}\mid \bar{\mathbf{z}}_d,\boldsymbol\eta_{l},y_{\mathrm{pa}(l),d}}\nonumber \\&\sim&\begin{cases}
\mathcal{N}(\bar{\mathbf{z}}^{T}_d\boldsymbol\eta_{l},1), & y_{\mathrm{pa}(l),d}=1 \\
\mathcal{N}(\bar{\mathbf{z}}^{T}_d\boldsymbol\eta_{l},1)\mathbb{I}(a_{l,d}<0), & y_{\mathrm{pa}(l),d}=-1 \end{cases}\label{zero_0}\end{eqnarray} %\item Draw $a_{l, d} \ | \ z_{1:N_d,d}, \boldsymbol\beta_l \sim \mathcal{N} \left(\bar z_d^{T} \boldsymbol\beta_{l},1\right)$, where $\bar z_d=N_d^{-1}\sum_{n=1}^{N_d}z_{n,d}$

\item Apply label $l$ to document $d$ according to $a_{l,d}$ \begin{eqnarray}\hspace{-2.2cm}
y_{l,d}\mid a_{l,d}=\begin{cases}
\;\;\, 1 & \text{if \ensuremath{a_{l,d}>0}}  \\% and \ensuremath{y_{{\rm \mathrm{pa}}(l),d}=1}}\\
 -1 & \text{otherwise} \end{cases}\label{zero_1}\end{eqnarray}

\end{itemize}
\end{itemize}
\end{enumerate}
\end{shortbox}

@@

The second half of step 4 is what is referred to as supervision in the supervised LDA literature.  This is where the hierarchical classification of the bag-of-words data takes place and the is-a label constraints are enforced.  For every label $l \in \mathcal{L}$, both the empirical topic distribution for document $d$ and whether or not its parent label was applied (i.e.~$\mathbb{I}(y_{\mathrm{pa}(l),d}=1)$) are used to determine whether or not  label $l$ is to be applied to  document $d$ as well.  Note that $y_{l,d}$ can only be applied to document $d$  (set to 1) if its parent label $\mathrm{pa}(l)$ was also applied (these expressions are specific to is-a constraints but could be modified to accommodate different constraints between labels).  The regression coefficients $\boldsymbol\eta_l$ are independent a priori, however, the hierarchical coupling in this model induces a posteriori dependence.   The net effect of this conditional hierarchy of profit regressors is that label predictors deeper in the label hierarchy are able to focus on finding features that distinguish the members of that set, conditioned on the fact that the members are in the same set.  It should be clear that one can trivially restrict this hierarchy to a depth one hierarchy.  Also, one can nearly as easily make the conditional classification at each node multi-class rather than single-class if more than one label at each node is required.  In many cases, however, a binary indicator along with a deeper or more complex tree is sufficient.
%We believe this to be a significant source of the empirical label prediction improvement we observe experimentally.  We test this hypothesis in Section~\ref{sec:experiments}.

Note that the choice of variables $a_{l,d}$ and how they are distributed were  driven at least in part by posterior inference efficiency considerations.  In particular, choosing probit-style auxiliary variable distributions for the $a_{l,d}$'s  yields conditional posterior distributions for both the auxiliary variables \eqref{eqn:a_l_d} and the regression coefficients \eqref{eqn:regression_param_conditional} which are analytic.  This simplifies posterior inference substantially.  A review of probit regression can be found near the end of this chapter in Section \ref{sec:probit_review}.

@@

In the common case where no negative labels are observed (like the example applications we consider in Section~\ref{sec:experiments}), the model must be explicitly biased towards generating  negative labels in order to keep it from learning to assign positive labels to all documents.   This is a common problem in modeling unbalanced data.  To see how this model can achieve this we draw the reader's attention to the $\mu$ parameter and, to a lesser extent, the $\sigma$ parameter above.  Because $\bar{\mathbf{z}}_d$ is always positive, setting $\mu$ to a negative value results in a bias towards negative labelings, i.e.~for large negative values of $\mu$, all labels become a priori more likely to be negative ($y_{l,d}=-1$).  We explore the effect of $\mu$ on out-of-sample label prediction performance in Section~\ref{sec:experiments}.   In a very real way $\mu$ is a knob that can be adjusted both before inference to induce a broad array of out-of-sample performance characteristics that vary along classical axes like specificity, recall, and accuracy.  A similar but less principled solution can be effected by changing the decision boundary from $0$ in \eqref{zero_0} and \eqref{zero_1}.  This technique can be used to vary out-of-sample label bias after learning.
